\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-24 coqimpl.tex
Adding "accessing family components outside that family" into lang-design section Slight change on syntactic translation section.
took a pass on lang-design and the early-going of coqimpl

Despite possible difficulties for future maintenance, this approach has
various advantages: 1. it is the easiest and the most accessible way to
prototype as it relieves us the necessity of familarity of Coq base,
especially for the implementation of the module and functor; 2. we have
a clear definition of trusted-base---the whole Coq; 3. it is easy to
debug---we just need to check the translated commands; 4. it can be
well-incorporated with the existent tools like VSCoq; 5. it is more
accessible for the interested audience who can then easily adapt our
plugin and give a try---otherwise they have to download and re-build the
whole customized Coq; 6. it is more stable because the surface syntax of
Coq should be stable across different versions; 7. still, this plugin
can capture the key ingredients of implementing family polymorphism
inside Coq and act as a reference for guiding an appropriate
implementation of Family Polymorphism inside all sorts of proof
assistants.\YZ{Does the Coq implementation perform type checking to
prevent illegal uses fam poly?}\EDJreply{Yes. Because I define feature of fampoly as a direct translation into Coq feature.  The fampoly is like a shorthand into multiple Coq commands. \\
Do you find translation from my current metatheory into MLTT (without linkage) helpful? \\
Actually I am also not sure what this question is aiming towards. \\
Are you complaining there is no formal specification of my fampoly feature in paper and thus there is no proof my Coq surface syntax is related to metatheory? If so, I thought our Coq plugin is only advertising how powerful Fampoly is to solve expression problem in mechanized proving. I can decouple the relationship between our plugin and our metatheory if you find the gap is really there... And I can only claim metatheory is about a partial result on incorporating fampoly with dependent type, Coq plugin is using two example to illustrate how mechanized proving can get benefit and our Coq plugin is only inspired by our metatheory (and not claim any relationship). Doing so would require rearrange the metatheory section before this section. \\
But I don't think the gap is there at all because if you want all possible gaps to disappear, then the action of using debruijn is wrong because we need to prove the one using debruijn and the surface syntax that is not using debruijn are ``equivalent''. What's more, module is not sigma type because Module is not a term in Coq and thus there is no formalism of Module at all. Let alone I don't see the plausibility of mechanizing dependent type without debruijn. \\ 
Please comment back so that I know what corresponding modification is needed on the paper. 
}

\textbf{Family compiled into module.} This is the main idea of compiling
family components into Coq's primitives: we compile family into Coq
module, family type into module type, and the context of a given term
into parameters of the module. Unlike module, we need to faithfully
reflect the \textit{late binding} nature of family: for each defining
field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using
``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall
\cL \sigma. T$. However, instead of using Coq's universal quantifier, we
choose module and module parameters to achieve this wrapping---each
family type in the context of the judgement will be one module parameter
with the compiled module type, and thus each field of the family will
compile into a functor parametrized by its context. Doing so we can also
get rapid feedback from type checking when the users are defining each
field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
Final Field a' : nat 
  := 1. 
Field b'
  : self_A.a' = 1
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}\YZ{Can the self's be omitted?}\EDJreply{Yes. But in our current plugin implementation it is not omitted. I think altering example is fine. If you think some small distance between example and the current implementation is fine, please comment back and I will remove all the self_}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user
interactively inputs the line "Field a : nat := ..", our plugin will
translate this statement into the "Module a_4", "Module Type a_5, a_6"
three components.\YZ{Typo? Are they a_4, a_5, a_6 in Figure 2?}\EDJreply{Thanks, fixed.}
{The "EmptySig" is just empty module type.} Then,
these modules and module types will be generated and type-checked by Coq
immediately. The computed "Module a_10" is then part of the context of
"Field b" (as the module type of "self__A") for "b" to refer to. The
internal representation of family "A" is simply the list of these
``compiled'' modules and module types. We can simply achieve incremental
checking and overriding as in \ref{langdesign-req0}---for example, if we
want to override "Field a", we simply replace that compiled module with
the new module inside the list. This compiled module type serves as an
abstraction that enables overriding.

@@

% Insert one pseudo-code example for explanation of the mechanism
% Use the natural number example
\begin{figure}[!htb]
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family B.
  FInductive b : Set 
    := tt : b | ff : b.
  Family neg_handler. 
    Field tt : self_B.b 
      := self_B.ff.
    Field ff : self_B.b 
      := self_B.tt.
  EndFamily.
FRecursor neg 
  about self_B.b 
  motive (fun _ => self_B.b)
  using self_B.neg_handler
  by _rec.
EndFamily.

>>>>

Despite possible difficulties for future maintenance, this approach has
various advantages: 1. it is the easiest and the most accessible way to
prototype as it relieves us the necessity of familarity of Coq base,
especially for the implementation of the module and functor; 2. we have
a clear definition of trusted-base---the whole Coq; 3. it is easy to
debug---we just need to check the translated commands; 4. it can be
well-incorporated with the existent tools like VSCoq; 5. it is more
accessible for the interested audience who can then easily adapt our
plugin and give a try---otherwise they have to download and re-build the
whole customized Coq; 6. it is more stable because the surface syntax of
Coq should be stable across different versions; 7. still, this plugin
can capture the key ingredients of implementing family polymorphism
inside Coq and act as a reference for guiding an appropriate
implementation of Family Polymorphism inside all sorts of proof
assistants.\YZ{Does the Coq implementation perform type checking to
prevent illegal uses fam poly?}\EDJreply{Yes. Because I define feature of fampoly as a direct translation into Coq feature.  The fampoly is like a shorthand into multiple Coq commands. \\
Do you find translation from my current metatheory into MLTT (without linkage) helpful? \\
Actually I am also not sure what this question is aiming towards. \\
Are you complaining there is no formal specification of my fampoly feature in paper and thus there is no proof my Coq surface syntax is related to metatheory? If so, I thought our Coq plugin is only advertising how powerful Fampoly is to solve expression problem in mechanized proving. I can decouple the relationship between our plugin and our metatheory if you find the gap is really there... And I can only claim metatheory is about a partial result on incorporating fampoly with dependent type, Coq plugin is using two example to illustrate how mechanized proving can get benefit and our Coq plugin is only inspired by our metatheory (and not claim any relationship). Doing so would require rearrange the metatheory section before this section. \\
But I don't think the gap is there at all because if you want all possible gaps to disappear, then the action of using debruijn is wrong because we need to prove the one using debruijn and the surface syntax that is not using debruijn are ``equivalent''. What's more, module is not sigma type because Module is not a term in Coq and thus there is no formalism of Module at all. Let alone I don't see the plausibility of mechanizing dependent type without debruijn. \\ 
Please comment back so that I know what corresponding modification is needed on the paper. 
}\YZreply{My original question was more directed towards the fact
that currently type checking is deferred to the Coq base. Generally,
when one uses a statically typed language as a compilation target, they
also want to have type checking at the source level, (1) because there
could be type errors at the source level that could not otherwise be
caught at the target level and (2) because it allows better error messages.
So it needs to be addressed why deferring type checking to Coq is OK.}

\textbf{Family compiled into module.} This is the main idea of compiling
family components into Coq's primitives: we compile families into Coq
modules, family types into module types, and the context of a given term
into parameters of the module. Unlike module, we need to faithfully
reflect the \textit{late binding} nature of families: for each defining
field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using
``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall
\cL \sigma. T$. However, instead of using Coq's universal quantifier, we
choose module and module parameters to achieve this wrapping---each
family type in the context of the judgement will be one module parameter
with the compiled module type, and thus each field of the family will
compile into a functor parametrized by its context. Doing so we can also
get rapid feedback from type checking when the users are defining each
field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
Final Field a' : nat 
  := 1. 
Field b'
  : self_A.a' = 1
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}\YZ{Can the self's be omitted?}\EDJreply{Yes. But in our current plugin implementation it is not omitted. I think altering example is fine. If you think some small distance between example and the current implementation is fine, please comment back and I will remove all the self_}\YZreply{If omitting the self's poses no technical challenges, then I'd say it's fine to also omit them in this figure.}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user
interactively inputs the line "Field a : nat := ..", our plugin will
translate this statement into the "Module a_4", "Module Type a_5, a_6"
three components.
{The "EmptySig" is just empty module type.} Then,
these modules and module types will be generated and type-checked by Coq
immediately. The compiled "Module a_10"\YZ{should this be a_4?} is then part of the context of
"Field b" (as the module type of "self_A") for "b" to refer to. The
internal representation of family "A" is the list of these
compiled modules and module types. We can achieve incremental
checking and overriding as in \ref{langdesign-req0}---for example, if we
want to override "Field a", we simply replace that compiled module with
the new module inside the list. This compiled module type serves as an
abstraction that enables overriding.

@@

% Insert one pseudo-code example for explanation of the mechanism
% Use the natural number example
\begin{figure}[!htb]
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
Family B.
  FInductive b : Set 
    := tt : b | ff : b.
  Family neg_handler. 
    Field tt : self_B.b 
      := self_B.ff.
    Field ff : self_B.b 
      := self_B.tt.
  EndFamily.
FRecursor neg 
  about self_B.b 
  motive (fun _ => self_B.b)
  using self_B.neg_handler
  by _rec.
EndFamily.

### 2022-06-24 lang-design.tex
took a pass on lang-design and the early-going of coqimpl

\YZ{looks more like a 'design space' rather than 'language design' section}\EDJreply{Good point. Do I just change the section name?s}
When adapting Family Polymorphism into dependent types, we choose to
focus only on the essence of family (and inheritance)
structure in \citet{zm2017}, and thus a lot of unrelated features, like
Interface, will be removed. In this case, it will look like module with
late-binding and inheritance as code and proof reuse for module. We also
need it to have good compatibility with inductive types, because we
don't want to lose too much reasoning power of Coq (and its tactic
programming).

@@

\begin{figure}[!htb]
  \begin{minipage}[t]{0.47\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family STLC.
Inductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
Inductive term : Set := 
  | var : id -> term 
  | lam : id -> term -> term ...
Fixpoint subst 
  : id -> term -> term := ...
Inductive has_type 
  : context -> term -> type := ...
Proposition subst_lemma :
  has_type (Γ; x ↦ A) t T ->
  has_type Γ u A ->
  has_type Γ (subst x u t) T.
Inductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family STLC_bool extends STLC.
Extend Inductive ty : Set :=
  | bool : ty.

@@

First and foremost, \textlabel{Req (0)}{langdesign-req0}\textbf{we desire incremental(modular) type checking}. Notice that, the modularity here in Coq is a bit different from other language, because Coq is \textit{interactive} theorem proving so we actually need \textit{statement-wise} incremental type checking, not only for saving the compilation time, but also for the immediate feedback and type-checking for the Coq developer.

@@

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction
reasoning, because the former one is \textit{proof-relevant}. This
difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:
\textbf{we need to reason about (the computation about) extensible
pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}.
When proving "subst_lemma", we have to know "subst" is invariant on
"tt", i.e., \mintinline{Coq}{(subst i x tt) = tt}. This kind of
information requires exposing \textit{computational rules} from the
recursors.\YZ{What goes wrong if subst is not invariant or you do not
know whether subst is invariant?}\EDJreply{Then it is unprovable because substitution lemma need to know this. You can just instantiate the whole lemma -- we want  
has_type (Γ; x ↦ A) tt T ->
has_type Γ u A ->
has_type Γ (subst i x tt) T. But this final goal requires to know the computation of subst.}

@@

\subsection{Language Design}
Inspired by our above requirements and Family Polymorphism, we propose the following extended syntax for Coq Vernacular Commands. 
% basically write down the specification of surface Coq commandline
%   and then explain  how the language design makes the programmer’s life easier
\begin{align*}
  V := \quad &"Family"~\langle ident\rangle~["extends"~\langle ident\rangle] 
  \quad |\quad "EndFamily" \\
  & |\quad "FRecursor"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle~"using"~\langle ident\rangle \\
  & |\quad "Extend"~"Family"~\langle ident\rangle \\ 
  & |\quad ["Extend"]~"FInductive"~\langle ident\rangle \\ 
  & |\quad ["Final" | "Override"]~"Field"~\langle ident\rangle~":"~\langle term \rangle~":="~\langle term \rangle \\
  & | \quad "Inherit"~\langle ident\rangle \quad | \quad "Inherit Until"~\langle ident\rangle
\end{align*}

We use "Family" and "EndFamily" to scope. Each field in one family is declared and defined by "Field". We can define nested family. Family will also be used to encode recursive handlers. Each case handling of one inductive type will be aggregated into one family, then "FRecursor" will be used to construct a recursor using these handlers. For example, \mintinline{Coq}{FRecursor recur about T motive f} "using" "handlers" will generate a new field "recur" that is doing induction (using elimination principle of "T") using motive "f", and the handlers are located in the family "handlers".

We can inherit a top level family "parent" using \mintinline{Coq}{Family ... extends parent} as a new top level family. As expected, "Override" will override/a field from the parent. "Inherit" can inherit nested family, inductive type and fields. The fields inside a family are order-sensitive because we have dependent type here. So "Inherit Until" is used if the user want to inherit a range of fields. If the user wants to extend an inductive type or a nested family from the parent, then "Extend FInductive" or "Extend Family" is used. Thus nested family can only be inherit or extended together with its outer family (and thus top level family), thus our support of family polymorphism is incomplete.

This incomplete support can already be helpful for extending functions. Take \cref{fig:plugin-example2} as example, if we want to extend boolean inductive type and its negation, we only need to specify the new constructor for boolean and how negation deal with the new constructor.

>>>>

When adapting Family Polymorphism into dependent types, we choose to
focus only on the essence of family (and inheritance)
structure in \citet{zm2017}, and thus a lot of unrelated features, like
Interface, will be removed. In this case, it will look like module with
late-binding and inheritance as code and proof reuse for module. We also
need it to have good compatibility with inductive types, because we
don't want to lose too much reasoning power of Coq\YZ{By 'reasoning power of Coq', I suppose you mean the expressive power of Gallina?}
(and its tactic programming).

@@

\begin{figure}[!htb]
  \begin{minipage}[t]{0.47\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=||]{Coq}
Family STLC.|\YZ{Does minted support adding extra keywords such as 'Family' and 'extends'?}|
Inductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
Inductive term : Set := 
  | var : id -> term 
  | lam : id -> term -> term ...
Fixpoint subst 
  : id -> term -> term := ...
Inductive has_type 
  : context -> term -> type := ...
Proposition subst_lemma :
  has_type (Γ; x ↦ A) t T ->
  has_type Γ u A ->
  has_type Γ (subst x u t) T.
Inductive step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family STLC_bool extends STLC.
Extend Inductive ty : Set :=
  | bool : ty.

@@

First and foremost, \textlabel{Req (0)}{langdesign-req0}: \textbf{we
desire incremental (modular) type checking}. Notice that, the modularity
here in Coq is a bit different from other languages, because Coq
supports \textit{interactive} theorem proving, so we actually need
\textit{statement-wise} incremental type checking, not only for avoiding
re\-compilation, but also for the immediate feedback and
type-checking for the Coq developer.

@@

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction
reasoning, because the former one is \textit{proof-relevant}. This
difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:
\textbf{we need to reason about (the computation about) extensible
pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}.
When proving "subst_lemma", we have to know "subst" is invariant on
"tt", i.e., \mintinline{Coq}{(subst i x tt) = tt}. This kind of
information requires exposing \textit{computational rules} from the
recursors.\YZ{What goes wrong if subst is not invariant or you do not
know whether subst is invariant?}\EDJreply{Then it is unprovable because substitution lemma need to know this. You can just instantiate the whole lemma -- we want  
has_type (Γ; x ↦ A) tt T ->
has_type Γ u A ->
has_type Γ (subst i x tt) T. But this final goal requires to know the
computation of subst.}\YZreply{Seems like what's missing here is to
convince the reader why exposing the definition of subst could be a
problem for family polymorphism.}

@@

\subsection{Language Design}
Inspired by our above requirements and Family Polymorphism, we propose
the following extension to Coq's Vernacular Commands.
% basically write down the specification of surface Coq commandline
%   and then explain  how the language design makes the programmer’s life easier
\begin{align*}
  V := \quad &"Family"~\langle ident\rangle~["extends"~\langle ident\rangle] 
  \quad |\quad "EndFamily" \\
  & |\quad "FRecursor"~\langle ident\rangle~"about"~\langle ident\rangle~"motive"~\langle term\rangle~"using"~\langle ident\rangle \\
  & |\quad "Extend"~"Family"~\langle ident\rangle \\ 
  & |\quad ["Extend"]~"FInductive"~\langle ident\rangle \\ 
  & |\quad ["Final" | "Override"]~"Field"~\langle ident\rangle~":"~\langle term \rangle~":="~\langle term \rangle \\
  & | \quad "Inherit"~\langle ident\rangle \quad | \quad "Inherit Until"~\langle ident\rangle
\end{align*}

We use "Family" and "EndFamily" to scope.
Each field in a family is declared and defined as a "Field".\YZ{This sentence sounds a bit recursive}
We can define nested families.
Family will also be used to encode recursive handlers. Each case
handling of one inductive type will be aggregated into one family, then
"FRecursor" will be used to construct a recursor using these handlers.
For example, \mintinline{Coq}{FRecursor recur about T motive f} "using"
"handlers" will generate a new field "recur" that is doing induction
(using elimination principle of "T") using motive "f", and the handlers
are located in the family "handlers".

We can inherit a top-level family "parent" using \mintinline{Coq}{Family
... extends parent} as a new top-level family. As expected, "Override"
will override a field from the parent. "Inherit" can be used to inherit
nested families, inductive types, and fields. The fields inside a family
are order-sensitive because we have dependent types here. So "Inherit
Until" is used if the user wants to inherit a range of fields. If the
user wants to extend an inductive type or a nested family from the
parent, then "Extend FInductive" or "Extend Family" is used. Thus nested
family can only be inherited or extended together with its outer family
(and thus top level family), thus our support of family polymorphism is
incomplete.

This incomplete support can already be useful for extending functions.
Take \cref{fig:plugin-example2} as example, if we want to extend boolean
inductive type and its negation, we only need to specify the new
constructor for boolean and how negation deals with the new constructor.
As a result, the programmer does not have to duplicate code---otherwise
the user will need to copy and paste the whole original negation
definition and add a new pattern matching case.
